Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-09-30 WSSSPE3/report/WSSSPE3_review.tex
use utf8
Added includes for every wg_main file
Former-commit-id: b07a1db0988ce04a85a5a6860db6a4c13a79e2d7
add names and titles of lightning talks
Former-commit-id: 5478422906ae768e27dceb886d50b047e87af0f7

\usepackage{color}
\usepackage{dcolumn}
\usepackage{float}
\usepackage{graphicx}
\usepackage[latin9]{inputenc}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{subfigure}
\usepackage{psfrag}
\usepackage{tabularx}
\usepackage[hyphens]{url}
\usepackage{wrapfig}
\usepackage{longtable}
\usepackage{verbatim}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Lightning Talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{need to update for WSSSPE3}
\begin{comment}
\note{Lead: Choi. Volunteers welcome.
\href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides.}}

Lighting talks, breakout groups, working groups are all different. Lightning
talks were new in WSSSPE2. They were offered to all paper contributors, and some
other attendees, those who got travel awards. Breakout groups in WSSSPE1 were
just discussions about topics; in WSSSPE2, the breakout groups were supposed to
come up with plans for actions in specific areas. The idea of working groups is
that they will come out of breakout groups as a way for the groups to actually
do the actions over time, not at the workshops.

Email addresses of speakers: gvwilson@software-carpentry.org, Colin Venters
<C.Venters@hud.ac.uk>, j.spencer@imperial.ac.uk, erinrobinson@esipfed.org,
marpierc@iu.edu, jwpeterson@gmail.com, abani@buffalo.edu, clenhardt@renci.org,
dsk@uchicago.edu, Samin.Ishtiaq@microsoft.com, james@howison.name,
s.a.harris@leeds.ac.uk, marcus.hanwell@kitware.com, thabermann@hdfgroup.org,
rdowns@ciesin.columbia.edu, cboettig@gmail.com, jakob.blomer@cern.ch,
editor@ascl.net
\end{comment}

Lightning talks were a new feature in WSSSPE2. Since the workshop program was
mostly dedicated to discussions, the organizers wanted to give the attendees a
chance to also `make a pitch' for an idea, either representing a contributed
paper or something different. Eighteen attendees volunteered to participate in
the lightning talks, each given only two minutes to speak and at most one slide
to show. The talks were presented in reverse alphabetic order of speakers' last
names. In the rest of this section, we highlight the gist of some of the
speakers' messages.

%\item Greg Wilson: Close Enough for Scientific Work

\begin{enumerate}
\item \textbf{Colin C. Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable
Software through Sustainable Software Architectures~\cite{Venters_poster}.}
Venters proposed that sustainable software is a composite, first-class,
non-functional requirement (NFR) that is at a minimum a measure of a system's
maintainability and extensibility, but may also include other NFRs such as
efficiency (e.g., energy, cost), interoperability, portability, reusability,
scalability, and usability. To achieve technically sustainable software, Venters
suggested that software architectures are fundamental as they are the primary
carrier of system NFRs, i.e., pre-system understanding; and influence how
developers are able to understand, analyze, extend, test, and maintain a
software system, i.e., post-deployment system understanding. In addition,
Venters highlighted that sustainability of software architectures needs to be
addressed to endure different types of change and evolution in order to mitigate
architectural drift, erosion, and knowledge vaporization.

%\item James Spencer: Developing New Developers

%\item Erin Robinson:  Sustaining Science Bazaars

\item \textbf{Marlon Pierce: Patching It Up, Pulling It
Forward~\cite{Pierce_poster}.} Pierce discussed how open open source is. Open
software needs a diverse, openly governed community behind it, just as it needs
open licensing and a public code repository. To probe the level of governance
within open source projects, he and his co-authors (Marru and Mattmann)
suggested a contest to encourage individual developers to submit
patches and requests to projects that are important to them. This simple
mechanism shall expose several governance mechanisms, such as how easy it is for
independent developers to communicate with project leadership, how projects
accept and license third-party contributions, and how projects make
decisions such as granting source tree write access.

\item \textbf{John Peterson: Continuous Integration for Concurrent MOOSE Framework
and Application Development on GitHub~\cite{Peterson_poster}.}
Peterson from the Idaho National Laboratory reported
that in March 2014, the MOOSE framework was released under an open source
license on GitHub, significantly expanding and diversifying the pool of current
active and potential future contributors on the project. The MOOSE team employs
an extensive continuous integration test suite to ensure that both the framework
and the applications based on the framework are verified before any code changes
are accepted into the repository. They use a combination of built-in Git
features such as branching and submodules, GitHub API integration capabilities,
and in-house developed testing software to perform this verification and update
the dependent applications in a relatively seamless manner for users.

\item \textbf{Abani Patra: Execute it~\cite{Patra_poster}.}
%Univ at Buffalo, SUNY
Patra discussed the value of an easily accessible platform for
executing scientific software, e.g., HUBzero to access XSEDE or other computing
resources. Such a platform for executing benchmark problems (even at a small
scale) allows the developer community access a reference implementation and
provides an easy way to train the larger user community. A second idea of this
talk was that for true usability, much more attention and support needs
to be given to the integrated use of simulation tools inside complex workflows.

%\item Christopher Lenhardt: Open Science for Synthesis (OSS): Filling the Gapin %Early Career Training

\item \textbf{Daniel S. Katz: Implementing Transitive Credit with
JSON-LD~\cite{Katz_transitive_credit_poster}.} Science and engineering research
increasingly relies on activities that facilitate research but are not rewarded
or recognized, such as: data sharing; developing common data resources, software
and methodologies; and annotating data and publications. To promote and advance
these activities, we must develop mechanisms for assigning credit, facilitate
the appropriate attribution of research outcomes, devise incentives for
activities that facilitate research, and allocate funds to maximize return on
investment. Katz discussed the issue of assigning credit for both direct and
indirect contributions by using JSON-LD to implement a prototype transitive
credit system.

\item \textbf{Samin Ishtiaq: Daemons, Notifications and Sustaining Software. }%~\cite{Ishtiaq_poster}}
%  Microsoft Research Cambridge
The reproduction and replication of novel results has become a major issue in
computer science, systems biology, and other computational disciplines. These
include both the inability to re-implement novel algorithms and approaches, and
lack of an agreement on how and what to benchmark these algorithms on. Ishtiaq
from Microsoft Research Cambridge pointed out these problems and made several
suggestions to address them.

\item \textbf{James Howison: Retract all Bit-Rotten Publications. }%~\cite{Howison_poster}
Howison sought to provoke discussion by proposing that papers whose workflows
are not kept current with the changing software ecosystem should be
automatically retracted. This would create an incentive for authors to keep
their software current and usable, rather than the current situation in which
every potential user has to do this individually. A softer version of the
proposal would identify papers whose software workflow has become bit-rotten and
allow others to keep the code up to date, either adding them as new authors of
the paper or providing credit for their academic service in some other form.

%\item Sarah Harris: Sustainability of Multidisciplinary Software: A Science
%Perspective

%\item Marcus D. Hanwell: We Need Tool Builders for Sustainable Scientific
%Software

%\item Ted Habermann: Communities as the Whole Product

\item \textbf{Robert Downs: Community Recommendations for Improving Sustainable
Scientific Software Practices~\cite{Downs_poster}.} Robert Downs, of the
Columbia University Center for International Earth Science Information Network
(CIESIN), described a focus group study conducted with the Science Software
Cluster (SSC) of the Federation of Earth Science Information Partners (ESIP).
For the study, almost 300 attendees of the 2014 Summer ESIP Meeting were invited
to participate in simultaneous roundtable discussions on sustainability of
science software. Over two-thirds of the roundtable focus groups responded to a
semi-structured survey that contained three sets of questions eliciting
recommendations for near-term actions of the community to improve sustainable
software practices. Initial analysis of the participants' responses to the
questionnaire revealed several suggestions, which included improving community
engagement and collaborative activities, increasing understanding and awareness,
and creating incentives to motivate sustainable science software practices. 
The ESIP SSC plans to engage the community in the recommended activities for
improving sustainable scientific software practices.

\item \textbf{Carl Boettiger: rOpenSci: Building Sustainable Software by
Fostering a Diverse Community~\cite{Boettiger_poster}.} Boettiger described how
the rOpenSci project has been successful by focusing not just on building
software but also on building a community of researchers who learn and adopt
their approaches to reproducible research and sustainable software practice.
Through outreach, mentoring, workshops, and hackathons, they have not only
reached new users, but also turned users into co-developers of robust software
and good practices to support data science research across a growing set of
disciplines.

\item \textbf{Jakob Blomer: The Need for a Versioned Data Analysis
Environment~\cite{Blomer_poster}.} Large-scale scientific endeavors, such as the
discovery of the Higgs boson at the Large Hadron Collider (LHC), often rely on
complex software stacks. Maintaining thousands of dependencies of software
libraries and operating system versions has shown that despite source code
availability, the setup and the validation of a minimal usable analysis
environment can easily become prohibitively expensive. In high-energy physics,
CernVM-FS, a special-purpose, open-source, versioning, and snapshotting file
system used to capture and distribute entire software stacks, proved to be
useful for providing instant access to ready-to-run data analysis environments.
% for the global scientific community as well as for
%interested citizens, for instance through the CERN Open Data Portal
%(http://opendata.cern.ch). \todo{revise}

\item \textbf{Alice Allen: Find it! Cite it! }%~\cite{Allen_poster}}
The Astrophysics Source Code Library (ASCL) is an online registry of
scientist-written software used in astronomy research. Their primary interest is
rendering research more transparent by making this software more discoverable for
examination. The ASCL is treated as a publication by an indexing resource for
astronomy, the Astrophysics Data System (ADS). ADS tracks citations to what it
indexes, including citations to ASCL entries. Increasing rewards for writing
software, whether through citation, transitive credit or other methods, gives
software authors a powerful reason to take the time to build sustainability into
their software and is an excellent way to drive community change.
\end{enumerate}

>>>>

\usepackage{color}
\usepackage{dcolumn}
\usepackage{float}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{subfigure}
\usepackage{psfrag}
\usepackage{tabularx}
\usepackage[hyphens]{url}
\usepackage{wrapfig}
\usepackage{longtable}
\usepackage{verbatim}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Lightning Talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{need to update for WSSSPE3}
\begin{comment}
\note{
\href{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/}{Slides.}}
\end{comment}

\begin{enumerate}
\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs. Stability:
Facilitating Research Training While Supporting Scientific Research}}

\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
Betz, Ruzanna Chitchyan, Letícia Duboc, Steve Easterbrook, Guillermo
Rodriguez-Navas and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
Sustainability}}

\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones
and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces to
Improve Productivity of Legacy Tools}}

\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins and Kaitlin
Thaney: \textit{Collaborative Software Development as Sustainable Software: Lessons
from Open Source}}

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure
for Geodynamics}}

\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications in the
Computational Infrastructure for Geodynamics}}

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain and Charles Vardeman: \textit{Data And Software Preservation for Open
Science (DASPOS)}}

\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods and
Susheel Varma: \textit{Research Software Engineering Groups in Universities: The Story
from the UK}}

\item \textbf{Dan Gunter, Sarah Poon and Lavanya Ramakrishnan: \textit{Bringing the
User into Building Sustainable Software for Science}}

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham and
Sergei Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
Biology: Lessons Learned on the KBase Project}}

\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson and Karan
Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
Scientists in Best Practices of Software Sharing}}

\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software Accreditation
Framework}}

\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering Support for
Sustainable Scientic Software}}

\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}

\item \textbf{Alice Allen: \textit{We’re giving away the store! (Merchandise not
included)}}

\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
Bo Zhang: \textit{Scientific Software Success: Developing Metrics While Developing
Community}}

\end{enumerate}
